Table

### 2021-09-01 manuscript_pnas.Rmd
edits: update results

Our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, any simplification of complex river networks may result in regional biodiversity loss via reduced environmental heterogeneity. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Our findings underscore the importance of preserving the structural complexity of ecosystems and support the modern conservation initiative that seeks to protect a diverse habitat portfolio [@schindlerPortfolioConceptEcology2015]. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and the spatial prioritization of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While the old "single large or several small" debate has led to a principle prioritizing a single large habitat [@diamondIslandDilemmaLessons1975], emerging perspectives posit the potential of coordinated networks of small habitats in protecting biodiversity [@fahrigWhySeveralSmall2020]. Understanding the role of branching complexity helps design such networks. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (mean annual air temperature [degree Celsius] and cumulative precipitation [mm]), and land use (the fractions of forest, urban, and agriculture, and dam density per watershed area [km^-2^]). We used Hydrologically Adjusted Elevations of MERIT Hydro [@yamazakiMERITHydroHighResolution2019] to delineate watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area), which were separated by either the ocean, large lakes/reservoirs ($\ge$ 10 km^2^ in the areal area), or large rivers ($\ge$ 5000 km^2^ in the watershed area). The ocean and large lakes/reservoirs were extracted from Global 1-second Water Body Map ver 1.0 [@yamazakiDevelopmentGlobal90m2015]. Watershed area was estimated based on the delineated watershed polygons. We estimated branching probability by fitting exponential distributions to histograms of branch length as $branch~length \sim Exp(\lambda)$ (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal). A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$.

We estimated the spatial averages of elevation, mean annual air temperature and cumulative precipitation for each watershed polygon, which were extracted from Hydrologically Adjusted Elevations of MERIT Hydro [@yamazakiMERITHydroHighResolution2019] and WorldClim 1.4 [@hijmansVeryHighResolution2005]. The fractions of forest, urban, and agriculture were calculated based on land use data in 2015 from Copernicus Global Land Service (100-m resolution) [@marcelbuchhornCopernicusGlobalLand2020]. The Global Reservoir and Dam Database, Version 1 (GRanDv1) [@lehnerGlobalReservoirDam2011] was used to estimate dam density per unit watershed area (km^-2^). We used ArcMap 10.7, SAGA 6.3.0, and the following R packages to perform geospatial analysis: 'RSAGA' [@brenningRSAGASAGAGeoprocessing2018], 'sf' [@pebesmaSimpleFeaturesStandardized2018] 'raster' [@hijmansRasterGeographicData2020] 'stars' [@pebesmaStarsSpatiotemporalArrays2020], and 'exactextractr' [@bastonExactextractrFastExtraction2020].

@@

We developed robust regression models with an M-estimation method to examine power laws of riverine biodiversity. The robust regression analysis is insensitive to outliers of a response variable; therefore, this method allows us to account for uncertainties of our diversity estimates (see **Fish community data**). Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $w$, $log_{10}~y_w$, were assumed to follow a normal distribution as $log_{10}~y_w \sim Normal(\mu_w, \sigma^2)$. The parameter $\mu_w$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~A$), log-transformed branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, the fraction of agricultural land use (logit-transformed), and dam density. Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control their effects statistically. We did not include the fractions of urban and forest land use in the statistical models because these variables were strongly correlated with the fraction of agricultural land use (**Figure S19**). Similarly, we did not use mean elevation in the statistical models because it showed a moderate correlation with air temperature (**Figure S19**). Correlations for other combinations of explanatory variables were weak to moderate.

>>>>

Our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, any simplification of complex river networks may result in regional biodiversity loss via reduced environmental heterogeneity. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Our findings underscore the importance of preserving the structural complexity of ecosystems and support the modern conservation initiative that seeks to protect a diverse habitat portfolio [@schindlerPortfolioConceptEcology2015a]. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and the spatial prioritization of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While the old "single large or several small" debate has led to a principle prioritizing a single large habitat [@diamondIslandDilemmaLessons1975], emerging perspectives posit the potential of coordinated networks of small habitats in protecting biodiversity [@fahrigWhySeveralSmall2020]. Understanding the role of branching complexity helps design such networks. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (mean annual air temperature [degree Celsius] and cumulative precipitation [mm]), and land use (the fractions of forest, urban, and agriculture, and dam density per watershed area [km^-2^]). We used MERIT Hydro [@yamazakiMERITHydroHighResolution2019] to delineate watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area), which were separated by either the ocean, large lakes/reservoirs ($\ge$ 10 km^2^ in the areal area), or large rivers ($\ge$ 5000 km^2^ in the watershed area). The ocean and large lakes/reservoirs were extracted from Global 1-second Water Body Map ver 1.0 [@yamazakiDevelopmentGlobal90m2015]. Watershed area was estimated based on the delineated watershed polygons. We estimated branching probability by fitting exponential distributions to histograms of branch length as $branch~length \sim Exp(\lambda)$ (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal). A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$.

We estimated the spatial averages of elevation, mean annual air temperature and cumulative precipitation for each watershed polygon, which were extracted from Hydrologically Adjusted Elevations of MERIT Hydro [@yamazakiMERITHydroHighResolution2019] and WorldClim 1.4 [@hijmansVeryHighResolution2005]. The fractions of forest, urban, and agriculture were calculated based on land use data in 2015 from Copernicus Global Land Service (100-m resolution) [@marcelbuchhornCopernicusGlobalLand2020]. The Global Reservoir and Dam Database, Version 1 (GRanDv1) [@lehnerGlobalReservoirDam2011] was used to estimate dam density per unit watershed area (km^-2^). We used ArcMap 10.7 and the following R packages to perform geospatial analysis: 'sf' [@pebesmaSimpleFeaturesStandardized2018] 'raster' [@hijmansRasterGeographicData2020] 'stars' [@pebesmaStarsSpatiotemporalArrays2020], and 'exactextractr' [@bastonExactextractrFastExtraction2020].

@@

We developed robust regression models with an M-estimation method to examine power laws of riverine biodiversity. The robust regression analysis is insensitive to outliers of a response variable; therefore, this method allows us to account for uncertainties of our diversity estimates (see **Fish community data**). Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $w$, $log_{10}~y_w$, were assumed to follow a normal distribution as $log_{10}~y_w \sim Normal(\mu_w, \sigma^2)$. The parameter $\mu_w$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~A$), log-transformed branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, the fraction of agricultural land use, and dam density. Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control their effects statistically. We did not include the fractions of urban and forest land use in the statistical models because these variables were strongly correlated with the fraction of agricultural land use (**Figure S19**). Similarly, we did not use mean elevation in the statistical models because it showed a moderate correlation with air temperature (**Figure S19**). Correlations for other combinations of explanatory variables were weak to moderate.

### 2021-09-01 supplementary_information_pnas.Rmd
edits: update results

**Figure S19** Correlation structure of potential explanatory variables for riverine diversity metrics. Numeric values in each cell are the Pearson's correlation coefficients. Positive and negative correlations were colored in blue and red, respectively, and darker colors indicate stronger correlations. Environmental variables (temperature, precipitation, elevation, logit fraction of forest, logit fraction of urban, logit fraction of agriculture, and dam density) were expressed as deviations from the regional averages to remove any regional effects.

>>>>

**Figure S19** Correlation structure of potential explanatory variables for riverine diversity metrics. Numeric values in each cell are the Pearson's correlation coefficients. Positive and negative correlations were colored in blue and red, respectively, and darker colors indicate stronger correlations. Environmental variables (temperature, precipitation, elevation, fraction of forest, fraction of urban, fraction of agriculture, and dam density) were expressed as deviations from the regional averages to remove any regional effects.

### 2021-09-01 table_kable.Rmd
edits: minor correction for table 1
edits: update results

Effect <- c("Intercept",
            "$log_{10}$ Watershed area", 
            "$log_{10}$ Branching probability",    
            "Region (Midwest vs. Hokkaido)",                             
            "Air temperature",                                         
            "Precipitation",
            "Logit fraction of agriculture",
            "Dam density")

@@

**Table 1** Estimated parameters of robust regression models explaining fish species richness in Hokkaido (Japan) and the Midwest (US). The standard errors (SE) are shown in parenthesis. Response variables were log-10 transformed. Climate and land use variables (air temperature, precipitation, logit fraction of agriculture, dam density) are deviations from the regional averages and were standardized to a mean of zero and a standard deviation of one prior to the analysis.

>>>>

Effect <- c("Intercept",
            "$log_{10}$ Watershed area", 
            "$log_{10}$ Branching probability",    
            "Region (Midwest vs. Hokkaido)",                             
            "Air temperature",                                         
            "Precipitation",
            "Fraction of agriculture",
            "Dam density")

@@

**Table 1** Estimated parameters of robust regression models explaining fish species richness in Hokkaido (Japan) and the Midwest (US). The standard errors (SE) are shown in parenthesis. Response variables were log-10 transformed. Climate and land use variables (air temperature, precipitation, fraction of agriculture, dam density) are deviations from the regional averages and were standardized to a mean of zero and a standard deviation of one prior to the analysis.
